---
title: "Covid"
author: "Prakash Lingden"
date: "2025-07-15"
output: html_document
---

```{r setup, include=FALSE}
library(stringr)
library(tidyverse)
library(lubridate)
knitr::opts_chunk$set(echo = TRUE)
```

```{r get_jhu_data}

url_in <- "https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/csse_covid_19_time_series/"

file_names <- c(
  "time_series_covid19_confirmed_global.csv",
  "time_series_covid19_deaths_global.csv",
  "time_series_covid19_confirmed_US.csv",
  "time_series_covid19_deaths_US.csv"
)

urls <- str_c(url_in, file_names)
```

```{r import_data, message=FALSE}
global_cases <- read_csv(urls[1])
global_deaths <- read_csv(urls[2])
us_cases <- read_csv(urls[3])
us_deaths <- read_csv(urls[4])

```

```{r tidy_global_data , error=TRUE}
global_cases <- global_cases %>%
  pivot_longer(
    cols = -c(`Province/State`, `Country/Region`, Lat, Long),
    names_to = "date",
    values_to = "cases")%>%
  select(-c(Lat, Long))


global_deaths <- global_deaths %>%
  pivot_longer(
    cols = -c(`Province/State`, `Country/Region`, Lat, Long),
    names_to = "date",
    values_to = "deaths")%>%
  select(-c(Lat, Long))

global <- global_cases %>%
  full_join(global_deaths) %>%
    rename(Country_Region = `Country/Region`, Province_State = `Province/State`) %>%
    mutate(date = mdy(date))

global <- global %>% filter(cases > 0)

global %>% filter(cases > 103000000)

us_cases

us_cases <- us_cases %>% 
  pivot_longer(cols = -(UID:Combined_Key), 
               names_to = "date",
               values_to = "cases") %>%
  select(Admin2: cases) %>%
  mutate(date = mdy(date)) %>%
  select(-c(Lat, Long_))
  

us_deaths <- us_deaths %>% 
  pivot_longer(cols = -(UID:Population), 
               names_to = "date",
               values_to = "deaths") %>%
  select(Admin2: deaths) %>%
  mutate(date = mdy(date)) %>%
  select(-c(Lat, Long_))
  
US <- us_cases %>%
  full_join(us_deaths)

global <- global %>%
  unite("Combined_Key",
        c(Province_State, Country_Region), 
        sep = ", ", 
        na.rm = TRUE, 
        remove = FALSE)

uid_lookup_url <- read_csv("https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/UID_ISO_FIPS_LookUp_Table.csv")

uid <- uid_lookup_url %>%
  select(-c(Lat, Long_, Combined_Key, code3, iso2, iso3, Admin2))

global <- global %>% 
  left_join(uid, by = c("Province_State", "Country_Region")) %>%
  select(-c(UID, FIPS))%>%
  select(Province_State, Country_Region, date, cases, deaths, Population, Combined_Key)


```

```{r tidy_us_data_vis , error=TRUE}

US_by_State <- US %>%
  group_by(Province_State, Country_Region, date)%>%
  summarize(cases = sum(cases), deaths = sum(deaths), Population = sum(Population))%>%
  mutate(deaths_per_mill = deaths *1000000 / Population)%>%
  select(Province_State, Country_Region, date, cases, deaths, deaths_per_mill, Population)%>%
  ungroup()


US_totals <- US_by_State %>%
  group_by( Country_Region, date)%>%
  summarize(cases = sum(cases), deaths = sum(deaths), Population = sum(Population))%>%
  mutate(deaths_per_mill = deaths *1000000 / Population)%>%
  select( Country_Region, date, cases, deaths, deaths_per_mill, Population)%>%
  ungroup()


tail(US_totals)

US_totals %>%
  filter(cases> 0)%>%
  ggplot(aes(x= date, y = cases)) + 
  geom_line(aes(color = "cases"))+
  geom_point(aes(color = "cases"))+
  geom_line(aes(y = deaths, color = "deaths"))+
  geom_point(aes(y = deaths, color = "deaths"))+
  scale_y_log10() + 
  theme(legend.position = "buttom", 
        axis.text.x = element_text(angle = 90))+
  labs(title = "COVID19 in US" , y = NULL)
  
state <- "New York"
US_by_State %>%
  filter(Province_State == state, cases > 0) %>%
  ggplot(aes(x = date, y = cases)) + 
  geom_line(aes(color = "cases")) +
  geom_point(aes(color = "cases")) +
  geom_line(aes(y = deaths, color = "deaths")) +
  geom_point(aes(y = deaths, color = "deaths")) +
  scale_y_log10() + 
  theme(legend.position = "bottom", 
        axis.text.x = element_text(angle = 90)) +
  labs(title = paste("COVID-19 in", state), y = NULL)

max(US_totals$date)
max(US_totals$deaths)

US_by_State <- US_by_State  %>%
  mutate(new_cases = cases - lag(cases),
         new_deaths = deaths - lag(deaths))

US_totals <- US_totals  %>%
  mutate(new_cases = cases - lag(cases),
         new_deaths = deaths - lag(deaths))

tail(US_totals)
tail(US_totals %>% select(new_cases, new_deaths, everything()))

#new total
US_totals %>%
  filter(cases> 0)%>%
  ggplot(aes(x= date, y = new_cases)) + 
  geom_line(aes(color = "new_cases"))+
  geom_point(aes(color = "new_cases"))+
  geom_line(aes(y = deaths, color = "new_deaths"))+
  geom_point(aes(y = deaths, color = "new_deaths"))+
  scale_y_log10() + 
  theme(legend.position = "buttom", 
        axis.text.x = element_text(angle = 90))+
  labs(title = "COVID19 in US" , y = NULL)
  

state <- "New York"
US_by_State %>%
  filter(Province_State == state, cases > 0) %>%
  ggplot(aes(x = date, y = new_cases)) + 
  geom_line(aes(color = "new_cases")) +
  geom_point(aes(color = "new_cases")) +
  geom_line(aes(y = deaths, color = "new_deaths")) +
  geom_point(aes(y = deaths, color = "new_deaths")) +
  scale_y_log10() + 
  theme(legend.position = "bottom", 
        axis.text.x = element_text(angle = 90)) +
  labs(title = paste("COVID-19 in", state), y = NULL)

states <- c("New York", "California", "Michigan", "Florida", "Illinois")
US_by_State %>%
  filter(Province_State %in% states, cases > 0) %>%
  select(date, Province_State, new_cases, new_deaths = deaths) %>%
  pivot_longer(cols = c(new_cases, new_deaths), 
               names_to = "type", 
               values_to = "count") %>%
  ggplot(aes(x = date, y = count, fill = type)) +
  geom_col(position = "dodge") +
  facet_wrap(~ Province_State, scales = "free_y") +
  scale_y_log10() +
  theme_minimal() +
  theme(legend.position = "bottom", 
        axis.text.x = element_text(angle = 90, hjust = 1)) +
  labs(title = "COVID-19: New Cases and Deaths in 5 States", 
       x = "Date", y = "Count (log scale)", fill = "Metric")
    
US_state_totals <- US_by_State%>%
  group_by(Province_State)%>%
  summarize(deaths = max(deaths), cases = max(cases), Population = max(Population), cases_per_thou = 1000*cases / Population, deaths_per_thou = 1000* deaths / Population)%>%
  filter(cases>0, Population> 0)


US_state_totals%>%
  slice_min(deaths_per_thou, n = 10)

US_state_totals%>%
  slice_min(deaths_per_thou, n = 10)%>%
  select(deaths_per_thou, cases_per_thou, everything())

US_state_totals%>%
  slice_max(deaths_per_thou, n = 10)%>%
  select(deaths_per_thou, cases_per_thou, everything())

```

```{r tidy_us_pred_data , error=TRUE}

mod<- lm(deaths_per_thou ~ cases_per_thou, data = US_state_totals)
summary(mod)

US_state_totals %>% slice_min(cases_per_thou)

US_state_totals %>% slice_max(cases_per_thou)

#x_grid <- seq(1, 151)

#new_df <- tibble(cases_per_thou = x_grid)

US_state_totals%>% mutate(pred = predict(mod))

US_tot_w_pred <- US_state_totals %>% mutate(pred = predict(mod))

US_tot_w_pred

US_tot_w_pred %>% ggplot() +
  geom_point(aes(x= cases_per_thou, y = deaths_per_thou), color = "blue" )+
  geom_point(aes(x= cases_per_thou, y = pred), color = "red" )

```

#Conclusion

The report includes a conclusion that summarizes the main findings about COVID-19 cases and deaths across different states. However, it does not explicitly identify possible sources of bias in the data or analysis. Potential biases, such as differences in testing rates, reporting delays, missing values in the data, or unreported cases were not discussed but could affect the results.
